 

### 2021-04-26 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and North, 2009]. Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009]. During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

Under colder conditions, minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory. Across Europe [@zellweger_seasonal_2019; Defrenne et al. 2021] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings. Greater radiative heat loss in exposed areas results in lower $T_{min}$. However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].

@@

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [Drake et al. 2018] or under high $T_{air}$ and stomtal limitation, can cause leaf heat stress [Fauset et al. 2018].

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; *Koch et al. 1994*]. Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al. 2018].  However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low. This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [*Koch et al. 1994*; @fauset_differences_2018]. On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

@@

Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. Greater heat stress and $T_{air}$ in the lower canopy closer is also observed in Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987].  Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016]. In other dry conditions, inner canopy can be warmer than above canopy and air temperature due to greater relative humidity [Dietz et al. 2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies, caused by midday stomatal closure that reduces transpirational cooling [Slot et al. 2019]. This is seen in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].

In three forest types- US temperate deciduous, Puerto Rico tropical wet and tropical moist, $T_{leaf_{max}}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in the upper canopies of all three forests, with increasing $T_{leaf}$ - $T_{air}$  with height. $T_{leaf}$ - $T_{air}$ was observed to be greatest in temperate deciduous forest compared to tropical moist and wet. Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [Doughty and Gouden, 2008; @mauTemperateTropicalForest2018a]

@@

**Leaf thermoregulation largely depends on leaf traits. Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; Zweiniecki et al 2004]**

@@

Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [Zweineicki et al. 2004]. Degrees of direct to diffused light and consequental water demands shape sun vs. shade leaves and their necessary vascular architecture [Casal et al. 2011, Zweineicki et al. 2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. Therefore, exposed sun leaves are typically smaller in size and width, thicker, hairier and have greater lobation, and buffer biophysical conditions for shade leaves. Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [Sack et al. 2006; Chazdon et al. 1996, Poorter et al. 2000; @rozendaalPlasticityLeafTraits2006; Mathur et al. 2018]. Respective leaf traits also play a role in heat dissapation where exposed sun leaves are more efficient in dissapting heat than buffered shade leaves [Vogel].

@@

Within-individuals exposed canopies substanially buffer inner and lower canopies from light and biophysical conditions. Thus leaf traits in canopy positions vary accordingly [Casal et al. 2011]. Several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [Chen et al. 2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves. Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016]. However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)

@@

Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air. [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al. 1999; Niinemetes, 1998]. Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al. 2010]. On the other hand, leaf temperature increases with effective leaf width and leaf size [Fauset et al. 2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017].

@@

Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs. shade leaf response to heatflecks under available soil nitrogen content [Kitao et al. 2018]. Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014]. Experiemental increase in upper canopy temperature, shows that LMA and $N_{area}$ tend to decrease [Carter and Cavaleri, 2014] due to increase in respiration **?** [**REF**]. Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [Poorter et al. 2019]

@@

Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [Rosendaal et al. 2006; Thomas and Winner, 2002; Houter and Pons 2012]. Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [Thomas and Winner, 2002; Houter and Pons 2012]. Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures. Therefore, some species might more susceptible to heat stress at a vertical gradient than others [Fauset et al. 2018].

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [Kenzo et al. 2015; Sanches et al. 2009]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [Sanches et al. 2009; Zweinniecki et al. 2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [Sanches et al. 2009; Hernandez]. Similarly, max transpiration rate increases with canopy height [Kenzo et al. 2015]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [Kosugi et al. 2021].

@@

Leaf temperature beyond photosynthetic optimum temperature, declines photosynthesis [**REF**]. Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; Slot & Winter 2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g. Slot & Winter 2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [Niinemetes et al. 1999b; Niinemetes and Valladares, 2004]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [Slot et al. 2019; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; Miller et al. 2020], and no significant trend in temperate trees [@mau_temperate_2018].

@@

Typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al. 2017). Across forest species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  Slot et al 2020; O' sullivan et al. 2017] and varies with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; Slot et al. 2020]. Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$. On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [Slot et al. 2020]. However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins. Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower thermal safety margins [Perez and Feeley, 2020.]

@@

For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [Slot et al. 2019]. In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions. In both species photosynthetic heat tolerance increased from June to August [Knapp et al. 1994]

In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient. For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions (Curtis et al. 2019). Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [Curtis et al. 2019].

>>>>

The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009]. Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009]. During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

Under colder conditions, minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory. Across Europe [@zellweger_seasonal_2019; *Defrenne et al. 2021*] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings. Greater radiative heat loss in exposed areas results in lower $T_{min}$. However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].

@@

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [@cavaleriColdbloodedForestsWarming2020] or under high $T_{air}$ and stomtal limitation, can cause leaf heat stress [@fauset_differences_2018].

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al. 2018].  However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low. This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018]. On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

@@

Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. Greater heat stress and $T_{air}$ in the lower canopy closer is also observed in Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987].  Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016]. In other dry conditions, inner canopy can be warmer than above canopy and air temperature due to greater relative humidity [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies, caused by midday stomatal closure that reduces transpirational cooling [Slot et al. 2019]. This is seen in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].

In three forest types- US temperate deciduous, Puerto Rico tropical wet and tropical moist, $T_{leaf_{max}}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in the upper canopies of all three forests, with increasing $T_{leaf}$ - $T_{air}$  with height. $T_{leaf}$ - $T_{air}$ was observed to be greatest in temperate deciduous forest compared to tropical moist and wet. Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]

@@

**Leaf thermoregulation largely depends on leaf traits. Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

@@

Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004]. Degrees of direct to diffused light and consequental water demands shape sun vs. shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. Therefore, exposed sun leaves are typically smaller in size and width, thicker, hairier and have greater lobation, and buffer biophysical conditions for shade leaves. Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [@zwienieckiHydraulicLimitationsImposed2004; @rozendaalPlasticityLeafTraits2006; @mathurPhotosyntheticEfficiencySun2018a]. Respective leaf traits also play a role in heat dissapation where exposed sun leaves are more efficient in dissapting heat than buffered shade leaves [@vogelConvectiveCoolingLow1970].

@@

Within-individuals exposed canopies substanially buffer inner and lower canopies from light and biophysical conditions. Thus leaf traits in canopy positions vary accordingly [@casasSunShadeLeaves2011]. Several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [Chen et al. 2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves. Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016]. However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)

@@

Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air. [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al. 1999; @niinemetsAdjustmentFoliageStructure1998]. Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al. 2010]. On the other hand, leaf temperature increases with effective leaf width and leaf size [Fauset et al. 2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017].

@@

Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs. shade leaf response to heatflecks under available soil nitrogen content [@kitaoCanopyNitrogenDistribution2018]. Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014]. Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [@poorterMetaanalysisPlantResponses2019a]

@@

Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [@@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures. Therefore, some species might more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]. Similarly, max transpiration rate increases with canopy height [@kosugiSeasonalFluctuationsTemperature2006]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].

@@

Leaf temperature beyond photosynthetic optimum temperature, declines photosynthesis [**REF**]. Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and no significant trend in temperate trees [@mau_temperate_2018].

@@

Typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al. 2017). Across forest species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017] and varies with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020]. Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$. On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020]. However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margins. Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]

@@

For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [@slotPhotosyntheticHeatTolerance2019a]. In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions. In both species photosynthetic heat tolerance increased from June to August [Knapp et al. 1994]

In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient. For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions (Curtis et al. 2019). Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [@curtis_intracanopy_2019].
